### 2023-09-28 paper/paper.tex
Paper - add BF and GA descriptions and begin evaluation
Paper basic layout

\documentclass[11pt]{article}
    \title{\textbf{Minimum Array Partition}}
    \author{Daniil Grbić, Ivan Gogić}
    \date{\today}

    \addtolength{\topmargin}{-3cm}
    \addtolength{\textheight}{3cm}

    \usepackage{amsmath}

    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt}

\begin{document}

\maketitle
\thispagestyle{empty}

\section{Problem description}

    Given a $n \times n$ array $A$ of non-negative integers, and a positive integer $p$,  we need to choose $p - 1$ horizontal dividers $0 = h_0 < h_1 < \dots < h_p = n$ and $p - 1$ vertical dividers $0 = v_0 < v_1 < \dots < v_p = n$ partitioning $A$ into $p^2$ blocks, so as to minimize the following expression:
    $$
    \max_{\substack{1 \leq i \leq p \\ 1 \leq j \leq p}} \enspace
    \sum_{\substack{v_{i-1} \leq x \leq v_i \\ h_{j-1} \leq y \leq h_j}}
    A[x, y]
    $$

\section{Methods}

\subsection{Brute Force (BF)}

    Firstly, we write a brute force algorithm for the purpose of finding optimal solutions for small test cases, which we will use to evaluate our optimisation algorithms.

    BF generates all possible partitions of the matrix, and chooses the best one of them. This essentially means that BF will need an unreasonable amount of time to solve all but the smallest of testcases.

    The complexity of this approach is $O(n,p) = \binom{n}{p}^2 * p^2$, indeed $\binom{n}{p}^2$ is the total number of partitions, while for each of them we need to check $p^2$ blocks. We find the sum of each block in constant time, by utilising a partial sum DP matrix.

    \subsubsection{Cutting optimization}

    A small optimization of the previous approach involves abandoning some partitions early when it is clear that they will not yield a better solution than the current known best.

    Consider this: we know the sum $S$ of the entire matrix $M$, and we know there are $p^2$ blocks in total. According to the Pigeonhole principle, at least one of said blocks will have a sum greater or equal to $\frac{S}{p^2}$.

    Alternatively, if we have checked $m < p^2$ blocks and their total sum is $S_m < S$, then the maximum value of the remaining blocks will be greater or equal to $\frac{S-S_m}{m-p^2}$.

    If at any point this sum happens to become greater than the best solution we have found beforehand, we can safely skip checking the rest of the partition in question. In practice, this speeds up the algorithm 3 to 5 times.

\subsection{Genetic Algorithm (GA)}

    A genetic algorithm (GA) is a p-metaheuristic inspired by the process of natural selection. Genetic algorithms are commonly used to generate high-quality solutions to optimization and search problems by relying on biologically inspired operators such as mutation, crossover and selection.

    We use a nearly textbook GA implemenation. Nevertheless, given the specificity of the problem, crossover and mutation implementation deserve a closer look.

    \subsubsection{Crossover}

    Crossover of two partitions can be done by an arithmetical approach. For simplicity, we will explain the process on a one-dimensional partition. This can be directly applied to two dimensions for the problem in question.

    Consider two individuals, $A$ and $B$, with partitions $[1, 3, 9]$ and $[2, 5, 7]$. Their child $C$ will try to take a middle path in the hopes of bettering their fitness, by taking the average of respective divider pairs: $C = [\frac{1+2}{2}, \frac{3+5}{2}, \frac{9+7}{2}] = [1.5, 4, 8]$. Since dividors have to be whole numbers, we fix $C$ by either rounding the first element down, or rounding it up (chosen randomly).

    \subsubsection{Mutation}

    Mutation is achieved by trying to make small movements of individua dividers with a given probability to do so, considering there is an available place to move.

\subsection{Simulated Annealing (SA)}

\subsection{Variable Neighbourhood Search (VNS)}

\section{Method evaluation}

\subsection{Test generation}
We wrote a helper script for generating testcases. The script takes 3 command line arguments: $n$, $p$, and $type$; where $type$ refers to how the matrix $M$ is generated:
\begin{description}
    \addtolength{\itemindent}{0.80cm}
    \itemsep0em 
    \item[Ones] fills $M$ with ones, mainly used for debugging
    \item[Random] fills the matrix with numbers from $(1, n/2)$
    \item[Linear], $M_{i,j} = i + j + 1$
    \item[Squared], $M_{i,j} = i^2 + j^2 + 1$
\end{description}

\subsection{Results}

All 4 methods are evaluated on a set of 10 representative testcases. Each algorithm is run 5 times on each prolem insatnce, and we record the best, worst, and average results and execution times.

Let us begin with testcases representing the $random$ type. For this type of problems, we can see that SA finds close to optimal solutions at a fraction of the time:

\begin{verbatim}
Testcase: tests/random_n50_p3.in
  BRUTE FORCE
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=2.35       WORST=2.55       AVG=2.45      
  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.81       WORST=0.81       AVG=0.81      
  SIMULATED ANNEALING (SA)
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.13       WORST=0.14       AVG=0.14      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.47       WORST=0.48       AVG=0.47

Testcase: tests/random_n50_p6.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
    Results  : BEST=1043       WORST=1046       AVG=1045.40   
    Time (s) : BEST=1.62       WORST=1.68       AVG=1.65      
  SIMULATED ANNEALING (SA)
    Results  : BEST=1043       WORST=1046       AVG=1043.60   
    Time (s) : BEST=0.97       WORST=1.04       AVG=1.00      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
    Results  : BEST=1043       WORST=1046       AVG=1043.60   
    Time (s) : BEST=3.18       WORST=3.44       AVG=3.28

Testcase: tests/random_n100_p5.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=100, num_iters=500, elitism_size=20]
    Results  : BEST=11024      WORST=11066      AVG=11033.40  
    Time (s) : BEST=5.52       WORST=5.71       AVG=5.63      
  SIMULATED ANNEALING (SA)
    Results  : BEST=11024      WORST=11066      AVG=11037.20  
    Time (s) : BEST=0.60       WORST=0.66       AVG=0.64      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
    Results  : BEST=11024      WORST=11066      AVG=11049.20  
    Time (s) : BEST=9.29       WORST=9.43       AVG=9.39 
\end{verbatim}

Next, let us compare the results achieved on the $linear$ test type.

\begin{verbatim}
Testcase: tests/linear_n20_p4.in
  BRUTE FORCE
    Results  : BEST=602        WORST=602        AVG=602.00    
    Time (s) : BEST=4.71       WORST=5.17       AVG=4.99      
  GENETIC ALGORITHM (GA) [pop_size=20, num_iters=100, elitism_size=4]
    Results  : BEST=602        WORST=602        AVG=602.00    
    Time (s) : BEST=0.17       WORST=0.18       AVG=0.17      
  SIMULATED ANNEALING (SA)
    Results  : BEST=602        WORST=602        AVG=602.00    
    Time (s) : BEST=0.32       WORST=0.34       AVG=0.33      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=400]
    Results  : BEST=602        WORST=608        AVG=603.20    
    Time (s) : BEST=0.18       WORST=0.20       AVG=0.19

Testcase: tests/linear_n30_p4.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=30, num_iters=150, elitism_size=6]
    Results  : BEST=1920       WORST=1968       AVG=1958.40   
    Time (s) : BEST=0.37       WORST=0.39       AVG=0.38      
  SIMULATED ANNEALING (SA)
    Results  : BEST=1920       WORST=1920       AVG=1920.00   
    Time (s) : BEST=0.32       WORST=0.35       AVG=0.34      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=900]
    Results  : BEST=1920       WORST=1968       AVG=1939.20   
    Time (s) : BEST=0.43       WORST=0.44       AVG=0.43 
\end{verbatim}

We find the results for $squared$ test the most intersting.

\begin{verbatim}
Testcase: tests/squared_n40_p3.in
  BRUTE FORCE
    Results  : BEST=214662     WORST=214662     AVG=214662.00 
    Time (s) : BEST=1.60       WORST=1.72       AVG=1.65      
  GENETIC ALGORITHM (GA) [pop_size=40, num_iters=200, elitism_size=8]
    Results  : BEST=214662     WORST=215757     AVG=215538.00 
    Time (s) : BEST=0.52       WORST=0.53       AVG=0.52      
  SIMULATED ANNEALING (SA)
    Results  : BEST=214662     WORST=214662     AVG=214662.00 
    Time (s) : BEST=0.13       WORST=0.13       AVG=0.13      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=1600]
    Results  : BEST=214662     WORST=215757     AVG=215538.00 
    Time (s) : BEST=0.30       WORST=0.31       AVG=0.30

Testcase: tests/squared_n40_p4.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=40, num_iters=200, elitism_size=8]
      Results  : BEST=124440     WORST=131775     AVG=125907.00 
      Time (s) : BEST=0.66       WORST=0.69       AVG=0.68      
  SIMULATED ANNEALING (SA)
      Results  : BEST=124440     WORST=124440     AVG=124440.00 
      Time (s) : BEST=0.31       WORST=0.33       AVG=0.32      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=1600]
      Results  : BEST=124440     WORST=131775     AVG=125907.00 
      Time (s) : BEST=0.76       WORST=0.81       AVG=0.78

Testcase: tests/squared_n100_p10.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=100, num_iters=500, elitism_size=20]
    Results  : BEST=858130     WORST=984543     AVG=941758.40 
    Time (s) : BEST=14.10      WORST=14.69      AVG=14.47     
  SIMULATED ANNEALING (SA)
    Results  : BEST=903805     WORST=1149356    AVG=1030165.60
    Time (s) : BEST=4.56       WORST=4.85       AVG=4.69      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
    Results  : BEST=837551     WORST=928305     AVG=868049.20 
    Time (s) : BEST=28.35      WORST=33.75      AVG=30.67 
\end{verbatim}

\section{Failed ideas}

\section{Conclusion}

\section{References}

\end{document}

### 2023-09-28 paper/paper.tex
Paper - add BF and GA descriptions and begin evaluation
Paper basic layout

@@
         Time (s) : BEST=0.18       WORST=0.20       AVG=0.19    
-
-Testcase: tests/linear_n30_p4.in
-  BRUTE FORCE - skipping due to speed
-  GENETIC ALGORITHM (GA) [pop_size=30, num_iters=150, elitism_size=6]
-    Results  : BEST=1920       WORST=1968       AVG=1958.40   
-    Time (s) : BEST=0.37       WORST=0.39       AVG=0.38      
-  SIMULATED ANNEALING (SA)
-    Results  : BEST=1920       WORST=1920       AVG=1920.00   
-    Time (s) : BEST=0.32       WORST=0.35       AVG=0.34      
-  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=900]
-    Results  : BEST=1920       WORST=1968       AVG=1939.20   
-    Time (s) : BEST=0.43       WORST=0.44       AVG=0.43 
-\end{verbatim}
-
-We find the results for $squared$ test the most intersting.
-
-\begin{verbatim}
-Testcase: tests/squared_n40_p3.in
-  BRUTE FORCE
-    Results  : BEST=214662     WORST=214662     AVG=214662.00 
-    Time (s) : BEST=1.60       WORST=1.72       AVG=1.65      
-  GENETIC ALGORITHM (GA) [pop_size=40, num_iters=200, elitism_size=8]
-    Results  : BEST=214662     WORST=215757     AVG=215538.00 
-    Time (s) : BEST=0.52       WORST=0.53       AVG=0.52      
-  SIMULATED ANNEALING (SA)
-    Results  : BEST=214662     WORST=214662     AVG=214662.00 
-    Time (s) : BEST=0.13       WORST=0.13       AVG=0.13      
-  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=1600]
-    Results  : BEST=214662     WORST=215757     AVG=215538.00 
-    Time (s) : BEST=0.30       WORST=0.31       AVG=0.30
-
-Testcase: tests/squared_n40_p4.in
-  BRUTE FORCE - skipping due to speed
-  GENETIC ALGORITHM (GA) [pop_size=40, num_iters=200, elitism_size=8]
-      Results  : BEST=124440     WORST=131775     AVG=125907.00 
-      Time (s) : BEST=0.66       WORST=0.69       AVG=0.68      
-  SIMULATED ANNEALING (SA)
-      Results  : BEST=124440     WORST=124440     AVG=124440.00 
-      Time (s) : BEST=0.31       WORST=0.33       AVG=0.32      
-  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=1600]
-      Results  : BEST=124440     WORST=131775     AVG=125907.00 
-      Time (s) : BEST=0.76       WORST=0.81       AVG=0.78
-
-Testcase: tests/squared_n100_p10.in
-  BRUTE FORCE - skipping due to speed
-  GENETIC ALGORITHM (GA) [pop_size=100, num_iters=500, elitism_size=20]
-    Results  : BEST=858130     WORST=984543     AVG=941758.40 
-    Time (s) : BEST=14.10      WORST=14.69      AVG=14.47     
-  SIMULATED ANNEALING (SA)
-    Results  : BEST=903805     WORST=1149356    AVG=1030165.60
-    Time (s) : BEST=4.56       WORST=4.85       AVG=4.69      
-  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
-    Results  : BEST=837551     WORST=928305     AVG=868049.20 
-    Time (s) : BEST=28.35      WORST=33.75      AVG=30.67 
 \end{verbatim}

### 2023-09-29 minarrpar/common/solution.py
Method evaluation

@@ -16,10 +16,9 @@ class Solution:
     def __repr__(self):
         return (
                 f'Solution(\n' +
-                f'  n = {self.n}\n' +
-                f'  p = {self.p}\n' +
                 f'  h = {self.h}\n' +
                 f'  v = {self.v}\n' +
+                f'  result = {self.value()}\n' +
                 f')'
         )
 
@@ -37,7 +36,7 @@ class Solution:
                 v = (self.v[i], self.v[i + 1] - 1)
                 current = self.__block_value(h, v)
                 largest = max(largest, current)
-        return largest
+        return int(largest)
 
     def make_small_change(self):
         choices = []

### 2023-09-29 minarrpar/methods/brute_force.py

@@ -2,11 +2,12 @@ import random
 import numpy as np
 from random import shuffle
 from itertools import combinations, product
-from tqdm.notebook import tqdm, trange
+from tqdm.notebook import tqdm
 from minarrpar.common.instance import Instance
+from minarrpar.common.solution import Solution
 
 
-def bf(instance: Instance):
+def bf(instance: Instance, disabled_pbar=True):
     result = float('inf')
     result_h, result_v = [], []
     combs = list(combinations(range(1, instance.n), instance.p - 1))
@@ -17,8 +18,9 @@ def bf(instance: Instance):
     total_combinations = len(combs) ** 2
     total_sum = np.sum(instance.matrix)
     total_block_count = instance.p ** 2
+    evaluated_blocks = 0
 
-    for (comb_h, comb_v) in tqdm(product(combs_h, combs_v), total=total_combinations):
+    for (comb_h, comb_v) in tqdm(product(combs_h, combs_v), total=total_combinations, disable=disabled_pbar):
         curr = 0
         visited_block_count = 0
         visited_block_sum = 0
@@ -38,12 +40,18 @@ def bf(instance: Instance):
                 continue
             break
         else:
-            result = min(curr, result)
+            if result > curr:
+                result = curr
+                result_h, result_v = comb_h, comb_v
 
-        if result == curr:
-            result_h, result_v = comb_h, comb_v
+        evaluated_blocks += visited_block_count
 
-    print('--------------------')
-    print(f'h = {list(result_h)}')
-    print(f'v = {list(result_v)}')
-    print(f'result = {result}')
+    # percentage_cut = 100 * (1 - evaluated_blocks / (total_block_count * total_combinations))
+    # print(f'{percentage_cut:.2f}% cut')
+
+    # print('--------------------')
+    # print(f'h = {list(result_h)}')
+    # print(f'v = {list(result_v)}')
+    # print(f'result = {result}')
+
+    return Solution(instance, list(result_h), list(result_v))

### 2023-09-29 minarrpar/methods/genetic_algorithm.py

@@ -46,14 +46,14 @@ def crossover(parent1: Individual, parent2: Individual, child: Individual):
     child.v = np.array(list(map(lambda x: round(x + random.choice([-0.1, 0.1])), (parent1.v + parent2.v) / 2)))
 
 
-def ga(instance: Instance, pop_size, num_iters, tournament_size, mutation_prob, elitism_size=0, verbose=False):
+def ga(instance: Instance, pop_size, num_iters, tournament_size, mutation_prob, elitism_size=0, disabled_pbar=True):
     if elitism_size > 0 and (pop_size - elitism_size) % 2 == 1:
         elitism_size += 1
 
     population = [Individual(instance) for _ in range(pop_size)]
     new_population = [Individual(instance) for _ in range(pop_size)]
 
-    for iteration in tqdm(range(num_iters)):
+    for iteration in tqdm(range(num_iters), disable=disabled_pbar):
         if elitism_size > 0:
             population.sort(key=lambda x: x.fitness, reverse=True)
             new_population[:elitism_size] = deepcopy(population[:elitism_size])
@@ -72,19 +72,14 @@ def ga(instance: Instance, pop_size, num_iters, tournament_size, mutation_prob, 
 
         population = deepcopy(new_population)
 
-        # display progress
-        if verbose:
-            best = max(population, key=lambda x: x.fitness)
-            prob = mutation_prob * (num_iters - iteration/2) / num_iters
-            padding = len(str(num_iters))
-            print(f'{iteration:>{padding}} / {num_iters} : mutation_prob={prob:.2f} best_value={best.value()}')
-
     best_individual = max(population, key=lambda x: x.fitness)
     h = best_individual.h
     v = best_individual.v
     value = best_individual.value()
 
-    print('--------------------')
-    print(f'h = {list(h)}')
-    print(f'v = {list(v)}')
-    print(f'result = {value}')
+    # print('--------------------')
+    # print(f'h = {list(h)}')
+    # print(f'v = {list(v)}')
+    # print(f'result = {value}')
+
+    return Solution(instance, list(h), list(v))

### 2023-09-29 minarrpar/methods/s_metaheuristics.py

@@ -5,13 +5,13 @@ from minarrpar.common.instance import Instance
 from minarrpar.common.solution import Solution
 
 
-def sa(instance: Instance, num_iters):
+def sa(instance: Instance, num_iters, disabled_pbar=True):
     solution = Solution(instance)
     value = solution.value()
     best_solution = [deepcopy(solution.h), deepcopy(solution.v)]
     best_value = value
 
-    for i in tqdm(range(1, num_iters + 1)):
+    for i in tqdm(range(1, num_iters + 1), disable=disabled_pbar):
         new_choices = solution.make_small_change()
 
         for choice in new_choices:
@@ -42,10 +42,12 @@ def sa(instance: Instance, num_iters):
                     else:
                         solution.v[choice[1]] -= choice[2]
 
-    print('--------------------')
-    print(f'h = {list(best_solution[0])}')
-    print(f'v = {list(best_solution[1])}')
-    print(f'result = {best_value}')
+    # print('--------------------')
+    # print(f'h = {list(best_solution[0])}')
+    # print(f'v = {list(best_solution[1])}')
+    # print(f'result = {best_value}')
+
+    return Solution(instance, list(best_solution[0]), list(best_solution[1]))
 
 
 def shake(solution, choices, k):
@@ -67,11 +69,11 @@ def revert(solution, choices, chosen):
             solution.v[choices[resource][1]] -= choices[resource][2]
 
 
-def vns(instance: Instance, num_iters, k_max, move_prob):
+def vns(instance: Instance, num_iters, k_max, move_prob, disabled_pbar=True):
     solution = Solution(instance)
     value = solution.value()
 
-    for _ in tqdm(range(1, num_iters + 1)):
+    for _ in tqdm(range(1, num_iters + 1), disable=disabled_pbar):
         new_choices = solution.make_small_change()
         k_curr = min(k_max, len(new_choices))
 
@@ -85,7 +87,9 @@ def vns(instance: Instance, num_iters, k_max, move_prob):
             else:
                 revert(solution, new_choices, choice)
 
-    print('--------------------')
-    print(f'h = {list(solution.h)}')
-    print(f'v = {list(solution.v)}')
-    print(f'result = {value}')
+    # print('--------------------')
+    # print(f'h = {list(solution.h)}')
+    # print(f'v = {list(solution.v)}')
+    # print(f'result = {value}')
+
+    return solution

### 2023-09-29 paper/paper.tex
Papar - add SA and VNS descriptions, begin conclusion

--- 2023-09-28 paper/paper.tex
@@
     \usepackage{amsmath}
+    \usepackage{hyperref}
 
@@
 
+	Simulated Annealing is a probabilistic optimization method inspired by annealing in metallurgy.
+It's effective for combinatorial optimization problems with large search spaces.
+SA starts with an initial solution and uses a temperature schedule to accept worse solutions with a probability.
+\\
+	During the SA process, small changes are introduced through random movements of block borders, allowing for exploration of the solution space.
+
 \subsection{Variable Neighbourhood Search (VNS)}
+
+	Variable Neighborhood Search (VNS) is an optimization technique that iteratively explores different neighborhoods of solutions.
+It begins with an initial solution and systematically varies the neighborhood structure to search for better solutions.
+\\
+	During the VNS process, changes are introduced through random movements of (multiple) block borders.
 
@@
 
+\subsection{Linear Programming}
+
+	We initially had an idea to solve the problem using linear programming, as it seemed easy to define all the conditions that needed to hold true.
+The problem arose when defining the cost function, as it required block borders as matrix indices, and it appears that it's not possible to index the array with LP variables.
+
 \section{Conclusion}
+
+	In conclusion, the Minimum Array Partition problem has been established as NP-hard, making it challenging to find optimal solutions efficiently.
+While there exist polynomial algorithms that can provide approximations within a constant factor, their implementation complexity often outweighs the benefits (our approximations factor is better).
+\\
+	As demonstrated in our study, ... \\
+	These methods strike a balance between computational efficiency and solution quality, making them valuable tools for addressing the Minimum Array Partition problem.
 
@@
 
+\href{https://doi.org/10.1007/3-540-63165-8_216}{Efficient array partitioning} \\
+\href{https://doi.org/10.1007/BFb0030123}{On the complexity of the generalized block distribution}
+
 \end{document}

### 2023-09-29 paper/paper.tex
Finalize paper and clean up the repo

--- 2023-09-28 paper/paper.tex
@@
     \usepackage{amsmath}
+    \usepackage{hyperref}
+    \usepackage[margin=1.1in,footskip=0.25in]{geometry}
 
@@
 
+	Simulated Annealing is a probabilistic optimization method inspired by annealing in metallurgy.
+It's effective for combinatorial optimization problems with large search spaces.
+SA starts with an initial solution and uses a temperature schedule to accept worse solutions with a probability.
+
+	During the SA process, small changes are introduced through random movements of block borders, allowing for exploration of the solution space.
+
 \subsection{Variable Neighbourhood Search (VNS)}
+
+	Variable Neighborhood Search (VNS) is an optimization technique that iteratively explores different neighborhoods of solutions.
+It begins with an initial solution and systematically varies the neighborhood structure to search for better solutions.
+
+	During the VNS process, changes are introduced through random movements of (multiple) block borders.
 
@@
 
+For the purpose of evaluating our optimization algorithms, we created 3 sets (random, linear, and squared) of 4 testcases of different dimensions. The dimensions can be categorized as:
+\begin{description}
+  \addtolength{\itemindent}{0.80cm}
+  \itemsep0em 
+  \item[Tiny] $n=50$, $p=3$ (BF friendy)
+  \item[Small] $n=80$, $p=6$
+  \item[Medium] $n=100$, $p=4$
+  \item[Large] $n=500$, $p=10$
+\end{description}
+
+Each optimization algorithm is run 5 times for every testcase, and we record the best, worst, and average results and execution times.
+
+We only test BF on tiny testcases to compare the results from optimization algorithms. We run BF once per testcase, since there is no randomness involved.
+
 \subsection{Results}
 
-All 4 methods are evaluated on a set of 10 representative testcases. Each algorithm is run 5 times on each prolem insatnce, and we record the best, worst, and average results and execution times.
-
-Let us begin with testcases representing the $random$ type. For this type of problems, we can see that SA finds close to optimal solutions at a fraction of the time:
+Let us begin with testcases representing the $random$ type:
 
@@
     BRUTE FORCE
+      Result   : 3769
+      Time (s) : 2.46      
+    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
       Results  : BEST=3769       WORST=3769       AVG=3769.00   
-    Time (s) : BEST=2.35       WORST=2.55       AVG=2.45      
-  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
-    Results  : BEST=3769       WORST=3769       AVG=3769.00   
-    Time (s) : BEST=0.81       WORST=0.81       AVG=0.81      
+      Time (s) : BEST=2.11       WORST=2.28       AVG=2.20      
     SIMULATED ANNEALING (SA)
       Results  : BEST=3769       WORST=3769       AVG=3769.00   
-    Time (s) : BEST=0.13       WORST=0.14       AVG=0.14      
+      Time (s) : BEST=0.28       WORST=0.29       AVG=0.28      
     VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
       Results  : BEST=3769       WORST=3769       AVG=3769.00   
-    Time (s) : BEST=0.47       WORST=0.48       AVG=0.47
+      Time (s) : BEST=0.46       WORST=0.48       AVG=0.47
 
-Testcase: tests/random_n50_p6.in
+  Testcase: tests/random_n80_p6.in
     BRUTE FORCE - skipping due to speed
-  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
-    Results  : BEST=1043       WORST=1046       AVG=1045.40   
-    Time (s) : BEST=1.62       WORST=1.68       AVG=1.65      
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=3964       WORST=4072       AVG=3985.60   
+      Time (s) : BEST=7.86       WORST=8.07       AVG=7.94      
     SIMULATED ANNEALING (SA)
-    Results  : BEST=1043       WORST=1046       AVG=1043.60   
-    Time (s) : BEST=0.97       WORST=1.04       AVG=1.00      
-  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
-    Results  : BEST=1043       WORST=1046       AVG=1043.60   
-    Time (s) : BEST=3.18       WORST=3.44       AVG=3.28
+      Results  : BEST=3964       WORST=3964       AVG=3964.00   
+      Time (s) : BEST=2.18       WORST=2.27       AVG=2.21      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
+      Results  : BEST=3964       WORST=3964       AVG=3964.00   
+      Time (s) : BEST=8.26       WORST=8.45       AVG=8.35
 
-Testcase: tests/random_n100_p5.in
+  Testcase: tests/random_n100_p4.in
     BRUTE FORCE - skipping due to speed
-  GENETIC ALGORITHM (GA) [pop_size=100, num_iters=500, elitism_size=20]
-    Results  : BEST=11024      WORST=11066      AVG=11033.40  
-    Time (s) : BEST=5.52       WORST=5.71       AVG=5.63      
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=17061      WORST=17061      AVG=17061.00  
+      Time (s) : BEST=5.06       WORST=5.22       AVG=5.11      
     SIMULATED ANNEALING (SA)
-    Results  : BEST=11024      WORST=11066      AVG=11037.20  
-    Time (s) : BEST=0.60       WORST=0.66       AVG=0.64      
+      Results  : BEST=17061      WORST=17061      AVG=17061.00  
+      Time (s) : BEST=0.68       WORST=0.68       AVG=0.68      
     VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
-    Results  : BEST=11024      WORST=11066      AVG=11049.20  
-    Time (s) : BEST=9.29       WORST=9.43       AVG=9.39 
+      Results  : BEST=17061      WORST=17061      AVG=17061.00  
+      Time (s) : BEST=4.75       WORST=4.85       AVG=4.80
+
+  Testcase: tests/random_n500_p10.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=326181     WORST=338514     AVG=330243.80 
+      Time (s) : BEST=24.44      WORST=25.95      AVG=24.95     
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=360022     WORST=552450     AVG=424930.40 
+      Time (s) : BEST=9.21       WORST=10.39      AVG=9.89      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
+      Results  : BEST=323192     WORST=417608     AVG=362644.00 
+      Time (s) : BEST=24.83      WORST=28.67      AVG=26.29   
 \end{verbatim}
+
+For this first type of matrices, we can see that SA tends to find optimal (or close to optimal) solutions at a fraction of the time for tiny, small, and medium tests.
+However, SA falls behind on the large test case, displaying a big difference between its best (360022) and worst (552450) runs.
+
+Both GA and VNS seem clearly better than SA, however our genetic approach has the better average result across all 5 runs, indicating it is much less reliant on chance to find a good solution.
 
@@
 \begin{verbatim}
-Testcase: tests/linear_n20_p4.in
+  Testcase: tests/linear_n50_p3.in
     BRUTE FORCE
-        Results  : BEST=602        WORST=602        AVG=602.00    
-        Time (s) : BEST=4.71       WORST=5.17       AVG=4.99      
-    GENETIC ALGORITHM (GA) [pop_size=20, num_iters=100, elitism_size=4]
-        Results  : BEST=602        WORST=602        AVG=602.00    
-        Time (s) : BEST=0.17       WORST=0.18       AVG=0.17      
+      Result   : 15080
+      Time (s) : 3.18      
+    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
+      Results  : BEST=15080      WORST=15318      AVG=15127.60  
+      Time (s) : BEST=2.15       WORST=2.24       AVG=2.19      
     SIMULATED ANNEALING (SA)
-        Results  : BEST=602        WORST=602        AVG=602.00    
-        Time (s) : BEST=0.32       WORST=0.34       AVG=0.33      
-    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=400]
-        Results  : BEST=602        WORST=608        AVG=603.20    
-        Time (s) : BEST=0.18       WORST=0.20       AVG=0.19    
+      Results  : BEST=15080      WORST=15080      AVG=15080.00  
+      Time (s) : BEST=0.28       WORST=0.30       AVG=0.29      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
+      Results  : BEST=15080      WORST=15318      AVG=15270.40  
+      Time (s) : BEST=0.40       WORST=0.46       AVG=0.42
+
+  Testcase: tests/linear_n80_p6.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=16055      WORST=16055      AVG=16055.00  
+      Time (s) : BEST=7.88       WORST=8.10       AVG=7.98      
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=15936      WORST=16055      AVG=15983.60  
+      Time (s) : BEST=2.23       WORST=2.42       AVG=2.33      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
+      Results  : BEST=16055      WORST=16055      AVG=16055.00  
+      Time (s) : BEST=8.13       WORST=8.32       AVG=8.24
+
+  Testcase: tests/linear_n100_p4.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=67760      WORST=68355      AVG=67879.00  
+      Time (s) : BEST=5.16       WORST=5.34       AVG=5.23      
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=67760      WORST=68355      AVG=67879.00  
+      Time (s) : BEST=0.71       WORST=0.71       AVG=0.71      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
+      Results  : BEST=67760      WORST=72500      AVG=69065.00  
+      Time (s) : BEST=4.03       WORST=4.87       AVG=4.38
+
+  Testcase: tests/linear_n500_p10.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=1371552    WORST=1621573    AVG=1454183.80
+      Time (s) : BEST=23.79      WORST=26.55      AVG=24.81     
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=1383025    WORST=2073344    AVG=1565182.40
+      Time (s) : BEST=9.15       WORST=11.14      AVG=9.76      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
+      Results  : BEST=1378620    WORST=2168270    AVG=1566251.40
+      Time (s) : BEST=21.42      WORST=29.76      AVG=25.80        
 \end{verbatim}
+
+Here the story is very similar to what we observed with the previous test type on smaller test cases.
+Once again SA is able to find a competitive solution in a fraction of the time.
+In fact, it even beats GA and VNS on the small test, abeit only by a small margin.
+
+Once again, SA has a big difference between its best and worst runs on the large test, similarly to VNS. GA displays the best performance, and the lowest dependance on chance.
+
+The last group of test cases belong to the $squared$ group:
+
+\begin{verbatim}
+  Testcase: tests/squared_n50_p3.in
+    BRUTE FORCE
+      Result   : 520720
+      Time (s) : 4.03      
+    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
+      Results  : BEST=520720     WORST=520720     AVG=520720.00 
+      Time (s) : BEST=2.13       WORST=2.22       AVG=2.17      
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=520720     WORST=520720     AVG=520720.00 
+      Time (s) : BEST=0.28       WORST=0.29       AVG=0.28      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
+      Results  : BEST=520720     WORST=525369     AVG=521649.80 
+      Time (s) : BEST=0.47       WORST=0.48       AVG=0.47
+
+  Testcase: tests/squared_n80_p6.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=902840     WORST=968898     AVG=916051.60 
+      Time (s) : BEST=8.06       WORST=8.19       AVG=8.12      
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=901215     WORST=922944     AVG=905560.80 
+      Time (s) : BEST=2.25       WORST=2.35       AVG=2.30      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
+      Results  : BEST=902840     WORST=902840     AVG=902840.00 
+      Time (s) : BEST=7.94       WORST=8.32       AVG=8.13
+
+  Testcase: tests/squared_n100_p4.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=4808345    WORST=4808345    AVG=4808345.00
+      Time (s) : BEST=5.10       WORST=5.16       AVG=5.14      
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=4807152    WORST=4807152    AVG=4807152.00
+      Time (s) : BEST=0.65       WORST=0.70       AVG=0.67      
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
+      Results  : BEST=4808345    WORST=5033479    AVG=4853371.80
+      Time (s) : BEST=4.16       WORST=4.76       AVG=4.61
+
+  Testcase: tests/squared_n500_p10.in
+    BRUTE FORCE - skipping due to speed
+    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
+      Results  : BEST=522518392  WORST=619537980  AVG=571390189.00
+      Time (s) : BEST=23.39      WORST=26.13      AVG=25.28     
+    SIMULATED ANNEALING (SA)
+      Results  : BEST=533205270  WORST=1038957839 AVG=680165623.40
+      Time (s) : BEST=9.39       WORST=11.87      AVG=10.50     
+    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
+      Results  : BEST=551863221  WORST=851774417  AVG=660668487.00
+      Time (s) : BEST=19.00      WORST=25.25      AVG=22.11 
+\end{verbatim}
+
+We find the results for $squared$ test the most interesting.
+
+The results are very similar to the previous group when we look at the smaller test dimensions.
+
+Interestingly, GA had one particularly bad run on the small squared test (968898 compared to its 902840 best).
+Looking at the average (916051.60), we conclude that only 1 out of 5 runs exhibited such behavior.
+Despite displaying the smallest reliance on chance before, it is important to note that all optimization algorithms do still depend on random number generation, and as such can be prone to individual bad peformances.
+
+The largest test once again exhibits the results we have observer beforehand, with GA having the smallest result interval, and the lowest average by a significant margin.
 
@@
 
+\subsection{Linear Programming}
+
+	We initially had an idea to solve the problem using linear programming, as it seemed easy to define all the conditions that needed to hold true.
+The problem arose when defining the cost function, as it required block borders as matrix indices, and it appears that it's not possible to index the array with LP variables.
+
 \section{Conclusion}
+
+In conclusion, the Minimum Array Partition problem has been established as NP-hard, making it challenging to find optimal solutions efficiently.
+While there exist polynomial algorithms that can provide approximations within a constant factor, their implementation complexity often outweighs the benefits (our approximations factor is better).
+
+As demonstrated in our study, optimization algorithms can be successfully used to solve the Minimum Array Partition problem, with genetic approach displaying the most consistent results, followed closely by VNS.
+SA remains as a good alternative for when we need to find a decent solution quickly.
+
+These methods strike a balance between computational efficiency and solution quality, making them valuable tools for addressing the Minimum Array Partition problem.
 
@@
 
+\href{https://doi.org/10.1007/3-540-63165-8_216}{Efficient array partitioning - Sanjeev Khanna, S. Muthukrishnan \& Steven Skiena} \\
+\href{https://doi.org/10.1007/BFb0030123}{On the complexity of the generalized block distribution - Michelangelo Grigni \& Fredrik Manne}
+
 \end{document}

### 2023-09-29 paper/presentation.tex
Add presentation (in Serbian)

% Inbuilt themes in beamer
\documentclass{beamer}

% Theme choice:
\usetheme{CambridgeUS}

% Title page details: 
\title{Minimum Array Partition Optimization} 
\author{Daniil Grbić, Ivan Gogić}
\date{\today}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\setlength{\parskip}{12pt}

\begin{document}

\begin{frame}
    \titlepage 
\end{frame}

\section{Opis problema}
\begin{frame}{Opis problema}
    Data je matrica dimenzija $n \times n$ nenegativnih celih brojeva, i broj $p \in \mathbb{N}$.

    Potrebno je izabrati $p-1$ vertikalnih i $p-1$ horizontalnih pregrada koje particionišu matricu na $p^2$ blokova tako minimizujemo sledeći izraz:

    $$
    \max_{\substack{1 \leq i \leq p \\ 1 \leq j \leq p}} \enspace
    \sum_{\substack{v_{i-1} \leq x \leq v_i \\ h_{j-1} \leq y \leq h_j}}
    A[x, y]
    $$

    Odnosno treba da minimuzjemo najveću među sumama elemenata tih blokova.

\end{frame}

\section{Naše metode}
\subsection{Brute Force}
\begin{frame}{Brute Force (BF)}

    \setlength{\parskip}{18pt}

    Algoritam grube sile nam je potreban kako bi utvrdili ispravnost optimizacionih metoda na malim primerima.

    Gledamo sve moguće particije (ima ih $\binom{n}{p}^2$), a sume blokova efikasno računamo koristeći matricu parcijalnih suma (dinamičko programiranje).

    Ubrzanje 3 do 5 puta sečenjem loših rešenja posmatrajući sume proverenih blokova po Dirihleovom principu.

\end{frame}

\subsection{Genetski algoritam}
\begin{frame}[fragile]{Genetski Algoritam}

    \setlength{\parskip}{18pt}

    Genetski algoritam obećava dobre rezultate, ako možemo da smislimo kako da vršimo ukrštanje i mutaciju.

    Van tih funkcija koristimo "školski" genetski algoritam.

    Mutacija je jednostavna: za svaku pregradu u rešenju postoji verovatnoća da se u toku generacije pomeri levo ili desno ukoliko je to mesto prazno.

\end{frame}

\begin{frame}[fragile]{Genetski Algoritam - Ukrstanje}

    Koristimo aritmetički pristup:

    \begin{verbatim}
        roditelj A : 1 3 9     roditelj A : 1 2 3
        roditelj B : 2 5 7     roditelj B : 2 3 4
            dete 1 : 1 4 8         dete 1 : 1 2 3
            dete 2 : 2 4 8         dete 2 : 2 2 4
                                      ...   ...
    \end{verbatim}

    Nastaje problem kada se pregrade preklapaju - to rešavamo ignorisanjem. Rešenja kod kojih se pregrade preklapaju nisu optimalna, i nikad neće na kraju biti izabrana, ali mogu poslužiti kao dobar međukorak.

\end{frame}

\subsection{Simulirano kaljenje i VNS}
\begin{frame}[fragile]{Simulirano kaljenje i VNS}

    \setlength{\parskip}{18pt}

    Klasični algoritmi kaljenja i VNS.

    Kod algoritma kaljenja, male promene postižemo slučajnim pomeranjem pregrada, čime i istražujemo prostor rešenja.

    Kod VNS, promene postižemo slučajnim pomeranjem više pregrada odjednom.

\end{frame}

\section{Rezultati}
\subsection{Generisanje Testova}
\begin{frame}[fragile]{Rezultati - Generisanje Testova}

    \setlength{\parskip}{2pt}

    Testove generišemo po različitim pravilima, naime:
    \begin{description}
        \addtolength{\itemindent}{0.80cm}
        \itemsep0em 
        \item[Random] popunjava matricu brojevima iz $(1, n/2)$
        \item[Linear] gde je $M_{i,j} = i + j + 1$
        \item[Squared] gde je $M_{i,j} = i^2 + j^2 + 1$
    \end{description}

    Za svaki tip testa generišemo po 4 konkretna test primera, sledećih dimenzija:

    \begin{description}
        \addtolength{\itemindent}{0.80cm}
        \itemsep0em 
        \item[Tiny] $n=50$, $p=3$
        \item[Small] $n=80$, $p=6$
        \item[Medium] $n=100$, $p=4$
        \item[Large] $n=500$, $p=10$
    \end{description}

    Svaki test primer propuštamo kroz svaki algoritam po 5 puta, i beležimo najbolje, najgore, i srednje rezultate i vremena izvršavanja.

\end{frame}

\subsection{Random Testovi}
\begin{frame}[fragile]{Rezultati - Random testovi}

    \fontsize{8.2pt}{1pt}

    \begin{verbatim}
    Testcase: tests/random_n100_p4.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=17061      WORST=17061      AVG=17061.00  
        Time (s) : BEST=5.06       WORST=5.22       AVG=5.11      
      SIMULATED ANNEALING (SA)
        Results  : BEST=17061      WORST=17061      AVG=17061.00  
        Time (s) : BEST=0.68       WORST=0.68       AVG=0.68      
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=17061      WORST=17061      AVG=17061.00  
        Time (s) : BEST=4.75       WORST=4.85       AVG=4.80

    Testcase: tests/random_n500_p10.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=326181     WORST=338514     AVG=330243.80 
        Time (s) : BEST=24.44      WORST=25.95      AVG=24.95     
      SIMULATED ANNEALING (SA)
        Results  : BEST=360022     WORST=552450     AVG=424930.40 
        Time (s) : BEST=9.21       WORST=10.39      AVG=9.89      
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=323192     WORST=417608     AVG=362644.00 
        Time (s) : BEST=24.83      WORST=28.67      AVG=26.29   
    \end{verbatim}

\end{frame}

\subsection{Linear Testovi}
\begin{frame}[fragile]{Rezultati - Linear testovi}

    \fontsize{8.2pt}{1pt}

    \begin{verbatim}
    Testcase: tests/linear_n100_p4.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=67760      WORST=68355      AVG=67879.00  
        Time (s) : BEST=5.16       WORST=5.34       AVG=5.23      
      SIMULATED ANNEALING (SA)
        Results  : BEST=67760      WORST=68355      AVG=67879.00  
        Time (s) : BEST=0.71       WORST=0.71       AVG=0.71      
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=67760      WORST=72500      AVG=69065.00  
        Time (s) : BEST=4.03       WORST=4.87       AVG=4.38

    Testcase: tests/linear_n500_p10.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=1371552    WORST=1621573    AVG=1454183.80
        Time (s) : BEST=23.79      WORST=26.55      AVG=24.81     
      SIMULATED ANNEALING (SA)
        Results  : BEST=1383025    WORST=2073344    AVG=1565182.40
        Time (s) : BEST=9.15       WORST=11.14      AVG=9.76      
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=1378620    WORST=2168270    AVG=1566251.40
        Time (s) : BEST=21.42      WORST=29.76      AVG=25.80   
    \end{verbatim}

\end{frame}

\subsection{Squared Testovi}
\begin{frame}[fragile]{Rezultati - Squared testovi}

    \fontsize{8.2pt}{1pt}

    \begin{verbatim}
    Testcase: tests/squared_n100_p4.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=4808345    WORST=4808345    AVG=4808345.00
        Time (s) : BEST=5.10       WORST=5.16       AVG=5.14      
      SIMULATED ANNEALING (SA)
        Results  : BEST=4807152    WORST=4807152    AVG=4807152.00
        Time (s) : BEST=0.65       WORST=0.70       AVG=0.67      
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=4808345    WORST=5033479    AVG=4853371.80
        Time (s) : BEST=4.16       WORST=4.76       AVG=4.61

    Testcase: tests/squared_n500_p10.in
      GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
        Results  : BEST=522518392  WORST=619537980  AVG=571390189.00
        Time (s) : BEST=23.39      WORST=26.13      AVG=25.28     
      SIMULATED ANNEALING (SA)
        Results  : BEST=533205270  WORST=1038957839 AVG=680165623.40
        Time (s) : BEST=9.39       WORST=11.87      AVG=10.50     
      VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
        Results  : BEST=551863221  WORST=851774417  AVG=660668487.00
        Time (s) : BEST=19.00      WORST=25.25      AVG=22.11 
    \end{verbatim}

\end{frame}

\subsection{Zakljucak}
\begin{frame}[fragile]{Zakljucak}

    \fontsize{8.2pt}{1pt}

    Pokazali smo da je moguće uspešno koristiti optimizacione algoritme za rešavanje Minimum Array Partition problema.

    Genetiski algoritam daje najkonzistentnije rezultate, VNS je tu odmah iza njega.

    Simulirano kaljenje je dobra alternativa kada možemo da prihvatimo neoptimalno rešenje, a sprečava nas vreme.

\end{frame}

\end{document}

### 2023-09-29 paper/presentation.tex
Fix GA crossover, update documentation accordingly

@@
 
-    Nastaje problem kada se pregrade preklapaju - to rešavamo ignorisanjem.
-Rešenja kod kojih se pregrade preklapaju nisu optimalna, i nikad neće na kraju biti izabrana, ali mogu poslužiti kao dobar međukorak.
+    Nastaje problem kada se pregrade preklapaju - to rešavamo tako što problematiče prepreke premestimo na slučajno izabrana slobodna mesta. To garantuje korektnost i pored toga poboljšava rešenje.
 

### 2023-09-29 minarrpar/common/solution.py

@@ -53,3 +53,16 @@ class Solution:
 
         shuffle(choices)
         return choices
+
+    def fix(self):
+        unique = set(self.v)
+        duplicate = len(self.v) - len(unique)
+        choices = set(range(1, self.n)).difference(unique)
+        part = sorted(list(unique) + sample(list(choices), duplicate))
+        self.v = np.array(part)
+
+        unique = set(self.h)
+        duplicate = len(self.h) - len(unique)
+        choices = set(range(1, self.n)).difference(unique)
+        part = sorted(list(unique) + sample(list(choices), duplicate))
+        self.h = np.array(part)

### 2023-09-29 minarrpar/methods/genetic_algorithm.py

@@ -44,6 +44,7 @@ def mutation(child: Individual, mutation_prob):
 def crossover(parent1: Individual, parent2: Individual, child: Individual):
     child.h = np.array(list(map(lambda x: round(x + random.choice([-0.1, 0.1])), (parent1.h + parent2.h) / 2)))
     child.v = np.array(list(map(lambda x: round(x + random.choice([-0.1, 0.1])), (parent1.v + parent2.v) / 2)))
+    child.fix()
 
 
 def ga(instance: Instance, pop_size, num_iters, tournament_size, mutation_prob, elitism_size=0, disabled_pbar=True):
